
# Getting And Cleaning: Project From John Hopkins University Through Coursera


The objectif of this project is to acquire and process data to obtain a tidy data set ready for analysis.

---

## Getting and cleaning data provide a method to be followed: 
>* Acquiring Data 
>* Transforming for a specific need.
>* Providing a tidy dataset

## Pre-requiresist
Before doing any of this step, we need to setup the working environment including path and directory for the incoming database. 

---
###**I) Acquiring Data:**
All start by getting the dataset first. Our dataset is zipfile on the link bellow. https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip
* Download the zipfile using **download.file()** function.
`download.file(fileUrl, destfile="./dbase/samsgs.zip")`  is the command issued to download zipfile from the website. Here:
    * **fileUrl** is the actual link provided above.
    * **destfile** is the new path of the downloaded zipfile with a new name called *samsgs*.
* Unzip the zipfile that is intended to provide the dataset.
`unzip(zipfile="./dbase/samsgs.zip", exdir="./dbase")`. This will make work done.

* Check out what is inside of that generated folder caused by unzip().
we need to provide new path and list all the things inside of that folder. A function **list.files()** which takes *reper* (directory of the new created folder) as argument: See command bellow
```{r eval=FALSE}
fich <- list.files(reper, recursive=TRUE)

[1] "activity_labels.txt"                         
[2] "features.txt"                                
[3] "features_info.txt"                           
[4] "README.txt"                                  
[5] "test/Inertial Signals/body_acc_x_test.txt"   
[6] "test/Inertial Signals/body_acc_y_test.txt"   
[7] "test/Inertial Signals/body_acc_z_test.txt"   
[8] "test/Inertial Signals/body_gyro_x_test.txt"  
[9] "test/Inertial Signals/body_gyro_y_test.txt"  
[10] "test/Inertial Signals/body_gyro_z_test.txt"  
[11] "test/Inertial Signals/total_acc_x_test.txt"  
[12] "test/Inertial Signals/total_acc_y_test.txt"  
[13] "test/Inertial Signals/total_acc_z_test.txt"  
[14] "test/subject_test.txt"                       
[15] "test/X_test.txt"          
[16] "test/y_test.txt"                             
[17] "train/Inertial Signals/body_acc_x_train.txt" 
[18] "train/Inertial Signals/body_acc_y_train.txt" 
[19] "train/Inertial Signals/body_acc_z_train.txt" 
[20] "train/Inertial Signals/body_gyro_x_train.txt"
[21] "train/Inertial Signals/body_gyro_y_train.txt"
[22] "train/Inertial Signals/body_gyro_z_train.txt"
[23] "train/Inertial Signals/total_acc_x_train.txt"
[24] "train/Inertial Signals/total_acc_y_train.txt"
[25] "train/Inertial Signals/total_acc_z_train.txt"
[26] "train/subject_train.txt"                     
[27] "train/X_train.txt"                           
[28] "train/y_train.txt" 
```

----
###**II) Transform for specific needs:**
  >  One of the main goals of this stage is to merge (combine) specific data. Until now, we don't have a working dataset yet.
  
  **1) Merging:**
  
The project assignment requires us to **merge** the training and test sets to create a single dataset with.
We have three components that deal with traing and test datasets: They are **activities, features, subjects.**
  
   * Acquire activities based information on trainning and testing.

   ```{r eval=FALSE}
   activityTrainData <- read.table(file.path(reper, "train", "y_train.txt"), header=FALSE)
   
   activityTestData <- read.table(file.path(reper, "test", "y_test.txt"), header=FALSE)
   ```
   
   * Acquire features based information on trainning and testing.
   
   ```{r eval=FALSE}
   featureTrainData <- read.table(file.path(reper, "train", "X_train.txt"), header=FALSE)

   featureTestData <- read.table(file.path(reper, "test", "X_test.txt"), header=FALSE)
   ```

   * Acquire subjects based information on trainning and testing.
   
   ```{r eval=FALSE}
   subjectTrainData <- read.table(file.path(reper, "train", "subject_train.txt"), header=FALSE)

   subjectTestData <- read.table(file.path(reper, "test", "subject_test.txt"), header=FALSE)
   ```

Merging data requires concatenation, creating new variable names or manipulating with existing name.

*  Example of concatenation:
I used rbind() command to concatenate by row each variable data  into one variable data (activity data, subject data, and feature data). In order words a reduction of variables. See scripts bellow.

  Storing data for activities:
   ```{r eval=FALSE}
   actData <- rbind(activityTrainData, activityTestData)
   ```
  Storing data for subjects:
   ```{r eval=FALSE} 
   subjData <- rbind(subjectTrainData, subjectTestData)
   ``` 
  Storing data for features:
   ```{r eval=FALSE} 
   featData <- rbind(featureTrainData, featureTestData)
   ```

* Example of assigning names:
Giving names to created variables above is a required task of the project assignments.

   
Assign **activity** as a name for the combined training and test activities' data.
```{r eval=FALSE}
names(actData) <- c("activity")
```
Assign **subject** as a name to the combined training and test subjects' data.
```{r eval=FALSE}
names(subjData) <- c("subject")
```
We need to capture the original names under `features.txt` to assign to "featData".
   ``` {r eval=FALSE} 
   FeatureOldName <- read.table(file.path(reper, "features.txt"), head=FALSE)
   ```
   
This is going to read the **feature.txt** file and store it under **FeatureOldName**. `FeatureOldName$V2` is the actual field where those names are located.

`names(featData ) <- FeatureOldName$V2` setup the names.

Now we can create the dataset of the 3 variables  by using cbind() function.

```{r eval=FALSE}
DataV <- cbind(subjData, actData)
DataF <-cbind(featData, DataV)
```

**`DataF`** is the reference for the merging part of the assignment. This is the working dataset we have worked for.

**2) Extracting data:**

Based on the assigned task, not all the information on the working dataset are useful. Therefore, the project assignment asked to extract or sort data on related to mean and standard deviation of the measurement. 
I used the function grep() to find any name that matches or contains characters mean, std, mean(), std().

 ```{r eval=FALSE}
 findDescName <- FeatureOldName$V2[grep("mean\\(\\)|std\\(\\)", FeatureOldName$V2]
 ```
 This command gives all the expected names.
   
*Create a subset data frame based on the descriptive variables, subjectData, actData*
```{r eval=FALSE} 
selcName <- c(as.character(findDescName), "subject", "activity")
   tid1<-  DataF[selcName]
```
**tid1** represents the subset data from **DataF**(merged dataset).

**3) Labelling variables:**

* Uses descriptive activity names to name the activities in the data set.
   
   * Check the what names are under "activity_labels.txt"
   
   ```{r eval=FALSE}
   activityNames <- read.table(file.path(reper, "activity_labels.txt"),header = FALSE)
   ```
   * We first convert column activity as string characters
   
   ```{r eval=FALSE}
   tid1$activity <- as.character(tid1$activity)
   tid1$activity[tid1$activity == 1] <- "Walking"
   tid1$activity[tid1$activity == 2] <- "Walking Upstairs"
   tid1$activity[tid1$activity == 3] <- "Walking Downstairs"
   tid1$activity[tid1$activity == 4] <- "Sitting"
   tid1$activity[tid1$activity == 5] <- "Standing"
   tid1$activity[tid1$activity == 6] <- "Laying"
   ```
   * Appropriately labels the data set with descriptive variable names.

**gsub()** function is used to perform all matches (http://www.inside-r.org/r-doc/base/sub)
```{r eval=FALSE}
names(tid1)<-gsub("^t", "time", names(tid1))
names(tid1)<-gsub("^f", "frequency", names(tid1))
names(tid1)<-gsub("Acc", "Accelerometer", names(tid1))
names(tid1)<-gsub("Gyro", "Gyroscope", names(tid1))
names(tid1)<-gsub("Mag", "Magnitude", names(tid1))
names(tid1)<-gsub("BodyBody", "Body", names(tid1))
```
----

##**III) Providing a tidy data set:**

* First tidy data set:

After committing change in **tid1** we obtain a tidy data set called **tidyData1**.

```{r eval=FALSE}
tidyData1 <- tid1
```
* As a project task:  From the data set in step 4, creates a second, independent tidy data set with average of each variable for each activity and each subject.

```{r eval=FALSE}
library(data.table)

data.dt <- data.table(tidyData1)
## Create tidy data set
TidyData <-aggregate(.~subject + activity, data.dt, mean)
## make in order:
TidyData <-TidyData[order(TidyData$subject, TidyData$activity),]
```
Save the TidyData on my current directory. This will be final ouput file in txt format. 
```{r eval=FALSE}
write.table(TidyData, file = "Tidy2.txt", row.names = FALSE)
```
